---
title: "Ch6"
author: "Esteban Correa"
date: "01/26/2022"
output: html_document
---

```{r}
library(rethinking)
library(tidyverse)
library(dagitty)
memory.size()
rm(list = ls())
```



# Exercises

## Easy

X.

## Medium.

### 6M1. 

Modify the DAG on page 190 to include the variable V, an unobserved cause of C and Y: variables should you condition on now? C ← V→ Y. Reanalyze the DAG. How many paths connect X to Y? Which must be closed? Which variables should you condition on now?

Original DAG:

```{r}
ori_dag <- dagitty( "dag { 
U[unobserved]
A->U
A->C
U -> X
U->B
C->Y
C->B
X->Y
}")
coordinates( ori_dag ) <- list( x=c(X=0,Y=1,C=1,U=0,B=0.5,A=0.5) , y=c(X=1,Y=1,C=0,U=0,B=0.5,A=-0.5) )
drawdag( ori_dag ) 
```


Modified DAG:

```{r}
mod_dag <- dagitty( "dag { 
U[unobserved]
V[unobserved]
V->C
V->Y
A->U
A->C
U -> X
U->B
C->Y
C->B
X->Y
}")
coordinates(mod_dag ) <- list( x=c(X=0,Y=1,C=1,U=0,B=0.5,A=0.5,V=1.5) , y=c(X=1,Y=1,C=0,U=0,B=0.5,A=-0.5,V=0.5) )
drawdag( mod_dag ) 
```

We have four alternative paths:

A) X<-U<-A->C->Y
B) X<-U->B<-C->Y
C) X<-U<-A->C<-V->Y
D) X<-U->B<-C<-V->Y

A should be closed, conditioning on A, Not C because it is a collider of V so it will open the path.
B is already closed because collider at B
C is already closed because collider at C
D is already closed because collider at B

Let's check

```{r}

adjustmentSets( ori_dag , exposure="X" , outcome="Y" )
adjustmentSets( mod_dag , exposure="X" , outcome="Y" )
```



## Hard.

### 6H1. 
Use the Waffle House data, data(WaffleDivorce), to find the total causal influence of number of Waffle Houses on divorce rate. Justify your model or models with a causal graph

```{r}
data(WaffleDivorce)
d<-WaffleDivorce
dag_6h1 <- dagitty( "dag { A -> D 
A -> M -> D 
A <- S -> M 
S -> W -> D
}") 
coordinates(dag_6h1 ) <- list( x=c(W=0,D=1,A=1,S=0,M=0.5) , y=c(W=1,D=1,A=0,S=0,M=0.5) )
drawdag( dag_6h1 ) 

adjustmentSets( dag_6h1 , exposure="W" , outcome="D" )
```

```{r}
d$A<-scale(d$MedianAgeMarriage)
d$D<-scale(d$Divorce)
d$MR<-scale(d$Marriage)
d$W<-scale(d$WaffleHouses)


m6h1<-quap(flist=alist(
  D~dnorm(mu,sigma),#
  mu<-a+ba*A+bm*MR+bw*W,
  a~dnorm(0,0.2),
  ba~dnorm(0,0.5),
  bm~dnorm(0,0.5),
  bw~dnorm(0,0.5),
  sigma~dexp(1)
),data = d)
precis(m6h1)

#plotting Waffle houses with others  at their means 
W.seq=seq(-3,3,length.out=50)
# GS.seq=seq(-3,3,length.out=50)
#use link to compute mu for each sample from posterior and for each area in A.seq
mu <- link( m6h1 , data=data.frame(A=0,W=W.seq,MR=0))
# summarize the distribution of mu
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.95 )
# plot it all 
plot( D ~ W , data=d , col=rangi2,xlim=range(W.seq),
      ylab="Outcome: Divorce rate (Std)",xlab="Predictor=Waffle houses (Std)" ) 
mtext("Counterfactual of W, holding A=0, MR=0")
lines( W.seq , mu.mean , lwd=2 ) 
shade( mu.PI , W.seq )
```




```{r}

m6h1c<-quap(flist=alist(
  D~dnorm(mu,sigma),#
  mu<-alpha+bw*W,
  alpha~dnorm(0,.5),
  bw~dnorm(0,0.5),
  sigma~dexp(1)
),data = d)
precis(m6h1c)

d$southidx=d$South+1

m6h1b<-quap(flist=alist(
  D~dnorm(mu,sigma),#
  mu<-alpha[southidx]+bw*W,
  alpha[southidx]~dnorm(0,.5),
  bw~dnorm(0,0.5),
  sigma~dexp(1)
),data = d)
precis(m6h1b,depth=2 )

par(mfrow=c(1,3))
plot( precis( m6h1b , depth=2 , pars="alpha" ) , xlab="expected divorce rate (std)" )


#plotting Waffle houses with others  at their means 
W.seq=seq(-3,3,length.out=50)
# GS.seq=seq(-3,3,length.out=50)
#use link to compute mu for each sample from posterior and for each area in A.seq
mu <- link( m6h1b , data=data.frame(southidx=1,W=W.seq))
# summarize the distribution of mu
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.95 )
# plot it all 
plot( D ~ W , data=d , col=rangi2,xlim=range(W.seq),
      ylab="Outcome: Divorce rate (Std)",xlab="Predictor=Waffle houses (Std)" ) 
mtext("Counterfactual of W, holding South=NO")
lines( W.seq , mu.mean , lwd=2 ) 
shade( mu.PI , W.seq )

#plotting Waffle houses with others  at their means 
W.seq=seq(-3,3,length.out=50)
# GS.seq=seq(-3,3,length.out=50)
#use link to compute mu for each sample from posterior and for each area in A.seq
mu <- link( m6h1b , data=data.frame(southidx=2,W=W.seq))
# summarize the distribution of mu
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.95 )
# plot it all 
plot( D ~ W , data=d , col=rangi2,xlim=range(W.seq),
      ylab="Outcome: Divorce rate (Std)",xlab="Predictor=Waffle houses (Std)" ) 
mtext("Counterfactual of W, holding South=YES")
lines( W.seq , mu.mean , lwd=2 ) 
shade( mu.PI , W.seq )
```


After closing the path using S, there is no real causal influence of the number of waffle houses on the divorce rate. Waffle house is a confounder.


### 6H2. 

```{r}
impliedConditionalIndependencies(dag_6h1)
```

Age and waffles are independent when conditioning on South:

```{r}
m6h2a<-quap(flist=alist(
  A~dnorm(mu,sigma),#
  mu<-alpha[southidx]+bw*W,
  alpha[southidx]~dnorm(0,.5),
  bw~dnorm(0,0.5),
  sigma~dexp(1)
),data = d)
precis(m6h2a,depth=2 )

par(mfrow=c(1,2))
#plotting Waffle houses with others  at their means 
W.seq=seq(-3,3,length.out=50)
# GS.seq=seq(-3,3,length.out=50)
#use link to compute mu for each sample from posterior and for each area in A.seq
mu <- link( m6h2a , data=data.frame(southidx=1,W=W.seq))
# summarize the distribution of mu
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.95 )
# plot it all 
plot( A ~ W , data=d , col=rangi2,xlim=range(W.seq),
      ylab="Outcome: Age (Std)",xlab="Predictor=Waffle houses (Std)" ) 
mtext("Counterfactual of W, holding South=NO")
lines( W.seq , mu.mean , lwd=2 ) 
shade( mu.PI , W.seq )

mu <- link( m6h2a , data=data.frame(southidx=2,W=W.seq))
# summarize the distribution of mu
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.95 )
# plot it all 
plot( A ~ W , data=d , col=rangi2,xlim=range(W.seq),
      ylab="Outcome: Age (Std)",xlab="Predictor=Waffle houses (Std)" ) 
mtext("Counterfactual of W, holding South=Yes")
lines( W.seq , mu.mean , lwd=2 ) 
shade( mu.PI , W.seq )
```


Divorce and south are independent when conditioning on age, marriage, waffles:

```{r}
m6h2b<-quap(flist=alist(
  D~dnorm(mu,sigma),#
  mu<-alpha[southidx]+ba*A+bm*MR+bw*W,
  alpha[southidx]~dnorm(0,.5),
  ba~dnorm(0,0.5),
  bm~dnorm(0,0.5),
  bw~dnorm(0,0.5),
  sigma~dexp(1)
),data = d)
precis(m6h2b,depth=2 )

```

Yes, divorce is conditional independent of south.

Marriage and waffles when conditioning on south:

```{r}
m6h2c<-quap(flist=alist(
  MR~dnorm(mu,sigma),#
  mu<-alpha[southidx]+bw*W,
  alpha[southidx]~dnorm(0,.5),
  bw~dnorm(0,0.5),
  sigma~dexp(1)
),data = d)
precis(m6h2c,depth=2 )

par(mfrow=c(1,2))
#plotting Waffle houses with others  at their means 
W.seq=seq(-3,3,length.out=50)
# GS.seq=seq(-3,3,length.out=50)
#use link to compute mu for each sample from posterior and for each area in A.seq
mu <- link( m6h2c , data=data.frame(southidx=1,W=W.seq))
# summarize the distribution of mu
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.95 )
# plot it all 
plot( MR ~ W , data=d , col=rangi2,xlim=range(W.seq),
      ylab="Outcome: Marriage (Std)",xlab="Predictor=Waffle houses (Std)" ) 
mtext("Counterfactual of W, holding South=NO")
lines( W.seq , mu.mean , lwd=2 ) 
shade( mu.PI , W.seq )

mu <- link( m6h2c , data=data.frame(southidx=2,W=W.seq))
# summarize the distribution of mu
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.95 )
# plot it all 
plot( MR ~ W , data=d , col=rangi2,xlim=range(W.seq),
      ylab="Outcome: Marriage (Std)",xlab="Predictor=Waffle houses (Std)" ) 
mtext("Counterfactual of W, holding South=Yes")
lines( W.seq , mu.mean , lwd=2 ) 
shade( mu.PI , W.seq )
```










