---
title: "Ch9"
author: "Esteban Correa"
date: "06/01/2022"
output:
  pdf_document: default
  html_document: default
---

```{r}
knitr::opts_chunk$set(echo=T,warning = F,message = F)
```

```{r}
library(rethinking)
library(tidyverse)
library(dagitty)
memory.size()
rm(list = ls())
memory.limit(size=64000)
gc()

theme_set(theme_minimal())
```

# Chapter 9. Markov Chain Monte Carlo

## 9.1. Good King Markov and his island kingdom

Imagine 10 islands in a circular pattern ordered by population size, where 1 is the smallest, and 10 the biggest:

```{r}
numweeks=1e5
positions<-rep(0,numweeks)
currentIsland=10
for (i in 1:numweeks){
  
  positions[i]<-currentIsland
  # he flips a coin 
  
  nextIsland=currentIsland+sample(c(-1,1),size=1,replace=T) # takes a sample from c vector
  if(nextIsland<1)
    nextIsland<-10
  if(nextIsland>10)
    nextIsland<-1
  
  #move
  prob2move<-nextIsland/currentIsland
  # print(prob2move)
  if (runif(1)<prob2move){
    currentIsland<-nextIsland #move
  }else{
    currentIsland<-currentIsland #stay additional week
  }
  
}
#first 100 visits
plot( 1:200 , positions[1:200] )
plot( table( positions ) )
```


## 9.4. Easy HMC:ulam

```{r}
set.seed(94)
data(rugged) 
d <- rugged 
d$log_gdp <- log(d$rgdppc_2000) 
dd <- d[ complete.cases(d$rgdppc_2000) , ] 
dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp) 
dd$rugged_std <- dd$rugged / max(dd$rugged) 
dd$cid <- ifelse( dd$cont_africa==1 , 1 , 2 )

m8.3<-quap(alist(
  log_gdp_std~dnorm(mu,sigma),
  mu<-a[cid]+b[cid]*(rugged_std-mean(dd$rugged_std)),
  a[cid]~dnorm(1,0.1),
  b[cid]~dnorm(0,0.3),
  sigma~dexp(1)
),data = dd)
precis(m8.3,depth = 2)

```

to run same gdp~ruggedness example into HMC we need to:
1. Convert all outcomes/variables before any process (compute logarithm at each chain interation does not make sense).
2. create a new data frame only including variables used in the model (helps to fight with NA values).

```{r}
dat_8.3<-list(
  log_gdp_std=dd$log_gdp_std,
  rugged_std=dd$rugged_std,
  rugged_mean=mean(dd$rugged_std),
  cid=as.integer(dd$cid)
)
str(dat_8.3)
```

Using lists instead of dataframes allow to have required variables with arbitrary lengths, helpful in MLMs or constants.

```{r}
# we recommend running this is a fresh R session or restarting your current session
# install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
# install_cmdstan()

# Rule: one short chain to debug, for chais for verification and inference

m9.1<-ulam(alist(
  log_gdp_std~dnorm(mu,sigma),
  mu<-a[cid]+b[cid]*(rugged_std-rugged_mean),
  a[cid]~dnorm(1,0.1),
  b[cid]~dnorm(0,0.3),
  sigma~dexp(1)
),data = dat_8.3,chains =4,cores=4,iter = 2000,warmup = 500)
show(m9.1)
precis(m9.1,depth = 2)
traceplot(m9.1)
trankplot(m9.1)
pairs(m9.1)
```

You can see how chains are mixing well as well as converging to a distribution. The equivalent Stan code is:

```{r}
stancode(m9.1)
```


### 9.5.3. Taming a wild chain

```{r}
set.seed(953)
y<-c(-1,1)
m9.2<-ulam(
  alist(
    y~dnorm(mu,sigma),
    mu<-alpha,
    alpha~dnorm(0,1000),
    sigma~dexp(0.0001)
  ),data=list(y=y),chains=3,iter = 2000
)

precis(m9.2,depth = 2)
traceplot(m9.2)
trankplot(m9.2)
pairs(m9.2)
```

The solution comes from restricting flat priors:

```{r}
set.seed(953)
y<-c(-1,1)
m9.3<-ulam(
  alist(
    y~dnorm(mu,sigma),
    mu<-alpha,
    alpha~dnorm(1,10),
    sigma~dexp(1)
  ),data=list(y=y),chains=3,iter = 2000,warmup = 500
)

precis(m9.3,depth = 2)
traceplot(m9.3)
trankplot(m9.3)
pairs(m9.3)
```

the example above illustrates the folk theorem of statistical computing. When our bayesian model  samples slowly or clunky, usually is due to something as simple as having omitted one prior distribution. When you are having trouble fitting a model, it often indicates a bad model.

### 9.5.4. Non-identifiable parameters

Remember the problem of highly correlated predictors (the two legs problem).

```{r}
set.seed(954)
y<-rnorm(100,mean=0,sd=1)

m9.4<-ulam(
  alist(
    y~dnorm(mu,sigma),
    mu<-a1+a2,
    a1~dnorm(0,1000),
    a2~dnorm(0,1000),
    sigma~dexp(1)
  ),data=list(y=y),chains=3,cores=3,log_lik=TRUE
)
precis(m9.4)
traceplot(m9.4)
trankplot(m9.4)
```

Let's restrict the search:

```{r}
m9.5<-ulam(
  alist(
    y~dnorm(mu,sigma),
    mu<-a1+a2,
    a1~dnorm(0,10),
    a2~dnorm(0,10),
    sigma~dexp(1)
  ),data=list(y=y),chains=3,cores=3,log_lik=TRUE
)
precis(m9.5)
traceplot(m9.5)
trankplot(m9.5)
compare(m9.4,m9.5)
```






